Report for Qualifying Exam

### 2024-08-20 src/sections/introduction.tex
introduction changes

\section{Introduction} \label{sec:introduction}
Language is a dynamic, living entity that continually adapts to reflect the changing world around us.
As language evolves, words shift in meaning, acquire new connotations, or are repurposed for entirely new contexts.
These changes occur as society, technology, and culture change.
When words change meaning, there is often a delay before search engines and online platforms fully reflect these new meanings.
This creates a \emph{data void} — an absence of relevant, up-to-date information based on the current understanding of the word.
This void represents a gap between evolving language and existing online content.

\para{Research goal.}
Understanding how words evolve is crucial because it allows us to anticipate where data voids might occur.
By tracking linguistic changes, our goal is to predict when a word might create a gap in digital information which could lead to a potential data void.
To achieve this, we have selected following research papers as a foundation for our research.

@@

The concept of semantic change, as presented in these papers, refers to the change in the contextual usage of words over time.
This can be particularly useful for analyzing how words evolve and ensuring that, as meanings shift, data voids are promptly recognized and addressed.

\para{Finding data voids is significantly important.}
Data voids have real-world implications, particularly in socio-political and sociotechnical domains.
When important terms, especially those tied to politics, social issues, or technology, develop new meanings,
data voids can be exploited to shape public perception.

Data voids, left unaddressed, can distort debates, influence elections, and shape collective beliefs, often in misleading or harmful directions.
Once a data void is identified, it’s critical to fill that gap with accurate, authoritative information.
Otherwise, bad actors can capitalize on the absence of reliable data to promote their own agenda.
By understanding linguistic changes, researchers and technology platforms can work together to prevent these voids from becoming filled with harmful or misleading content.

\para{Example.}
The term \emph{crisis actor} has undergone a significant shift in meaning,
exemplifying the phenomenon of semantic change we've been discussing.

Historically, `crisis actor' referred to actors employed in emergency preparedness training exercises.
These actors simulate victims or other roles in staged disaster drills to help first responders practice their response to crises like natural disasters or terrorist attacks.
It was a technical, niche term with a specific and legitimate use.

However, in recent years, it is used to describe the idea that victims of real-world tragedies are actually actors.
For example, in mass shootings, the victims have been referred to in propaganda and misinformation campaigns as `crisis actors,'
falsely implying that the events were staged or fabricated.

This example clearly demonstrates how a data void occurs when a word’s meaning is co-opted and evolves rapidly,
leaving a gap in reliable, accurate content.
This allows misinformation to flourish, having a profound impact on both political discourse and social trust.\\

This report introduces methodologies to identify shifts in words and this will be useful in my research to proactively identify data voids based on two key papers.
To contextualize our work, we present a thorough literature review examining the evolution of word meaning (\Cref{sec:relatedwork}).
Subsequently, Section \Cref{sec:paper_kulkarni} delves into the first paper, focusing on significant linguistic change detection,
while Section \Cref{sec:paper_hamilton} explores the second paper on diachronic word embeddings.

>>>>

\section{Introduction} \label{sec:introduction}
Language is a dynamic, living entity that continually adapts to reflect the changing world around us.
As language evolves, words shift in meaning, acquire new connotations, or are repurposed for entirely new contexts.
These changes occur as society, technology, and culture change.
In the digital age, search engines and online platforms play a crucial role in shaping access to knowledge and information.
When words change meaning, there is often a delay before search engines and online platforms fully reflect these new meanings.
This creates a \emph{data void} — an absence of relevant, up-to-date information based on the current understanding of the word.
This void represents a gap between evolving language and existing online content.

\para{Research goal.}
Understanding how words evolve is crucial because it allows us to anticipate where data voids might occur.
By tracking linguistic changes, we aim to predict when a word might create a gap in digital information which could lead to a potential data void.
To achieve this, we have selected the following research papers as a foundation for our study.

@@

These papers present the concept of semantic change, which refers to the evolution of a word’s contextual usage over time.
This approach is particularly useful for analyzing how words evolve and ensuring that, as meanings shift, data voids are promptly recognized and addressed.

\para{Finding data voids is significantly important.}
Data voids have real-world implications, particularly in socio-political and sociotechnical domains.
When important terms, especially those tied to politics, social issues, or technology, gain new meanings,
people can exploit data voids to shape public perception.

Data voids, left unaddressed, can distort debates, influence elections, and shape collective beliefs, often in misleading or harmful ways.
Once a data void is identified, it’s critical to fill that gap with accurate, authoritative information.
Otherwise, bad actors can take advantage of the absence of reliable data to promote their own agendas.
By understanding linguistic changes, researchers and technology platforms can work together to prevent harmful or misleading content from filling these voids.

\para{Example.}
The term \emph{crisis actor} has significantly shifted in meaning,
exemplifying the phenomenon of semantic change we've been discussing.

Historically, `crisis actor' referred to actors employed in emergency preparedness training exercises.
These actors simulated victims or other roles in staged disaster drills to help first responders practice their response to crises like natural disasters or terrorist attacks.
It was a technical, niche term with a specific and legitimate use.

However, in recent years, it is used to describe the false idea that victims of real-world tragedies are actually actors.
For example, in mass shootings, propaganda and misinformation campaigns have referred to victims as “crisis actors,” falsely implying that the events were staged or fabricated.

This example clearly demonstrates how a data void occurs when people co-opt a word’s meaning and cause it to evolve rapidly,
leaving a gap in reliable, accurate content.
This allows misinformation to flourish, having a profound impact on both political discourse and social trust.\\

This report introduces methodologies to identify shifts in words and this will be useful in our research to proactively identify data voids based on two key papers.
To contextualize our work, we provide a thorough literature review examining the evolution of word meaning (\Cref{sec:relatedwork}).
Subsequently, Section \Cref{sec:paper_kulkarni} delves into the first paper, focusing on significant linguistic change detection,
while Section \Cref{sec:paper_hamilton} explores the second paper on diachronic word embeddings.

### 2024-08-20 src/sections/paper_hamilton.tex
introduction changes

\para{Discovering Shifts from data.}
In this task, the authors aimed to see if the methods could discover reasonable shifts in word meanings without prior knowledge of those shifts.
They examined the top-10 words that changed the most from the 1900s to the 1990s (\Cref{tab:discovering-shifts-table}) based on semantic displacement metric (\Cref{subsec:measuring-semantic-change}).
These words were categorized as \emph{genuine, borderline, or artifacts of the corpus} by the authors.

@@

\begin{table}[tbh]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}}                                          & \multicolumn{3}{c}{\cellcolor[HTML]{FFFFFF}\textbf{Examples}}                                                                                                                                                                                                                               \\ \cmidrule(l){2-4}
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\multirow{-2}{*}{\cellcolor[HTML]{FFFFFF}\textbf{Semantic Shift}}} & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{PPMI}}                                                    & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SVD}}                  & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SGNS}}                                        \\ \midrule
\rowcolor[HTML]{EFEFEF}
\textbf{Genuine Shift}                                                                & started                                                                                                      & \begin{tabular}[c]{@{}l@{}}headed, calls,\\ gay, actually\end{tabular}    & \begin{tabular}[c]{@{}l@{}}wanting, gay, check, starting,\\ major, actually, headed\end{tabular} \\
\cellcolor[HTML]{FFFFFF}\textbf{Borderline}                                           & \begin{tabular}[c]{@{}l@{}}know, got, would, \\ decided, think, stop, \\ remember, must, wanted\end{tabular} & male, naturally                                                           & touching                                                                                         \\
\rowcolor[HTML]{EFEFEF}
\textbf{Corpus Artifact}                                                              & -                                                                                                             & \begin{tabular}[c]{@{}l@{}}harry, wherever,\\ special, cover\end{tabular} & harry, romance                                                                                   \\ \bottomrule
\end{tabular}
\caption{The top 10 words identified by each embedding method were examined, and after reviewing the literature and analyzing their nearest neighbors,
the authors categorized these words into different types of semantic shifts.
For instance, the word “headed” shifted from primarily meaning the “top of a body or entity” to signifying “a direction of travel” indicated a \emph{genuine} shift.
Additionally, some \emph{borderline} cases were identified, which were largely attributed to global shifts in genre or discourse, as well as corpus artifacts.
For example, words like “special,” “cover,” and “romance” appeared to be \emph{artifacts} from the covers of fiction books, occasionally influenced by advertisements,
taken from \cite{hamilton-etal-2016-diachronic}.}
\label{tab:discovering-shifts-table}
\end{table}

>>>>

\para{Discovering Shifts from data.}
In this task, the authors aimed to see if the methods could discover reasonable shifts in word meanings without prior knowledge of those shifts.
They examined the top-10 words that changed the most from the 1900s to the 1990s (\Cref{tab:discovering-shifts-table}) based on semantic displacement metric (\Cref{subsec:measuring-semantic-change}).
The authors categorize the words identified as having shifted meanings into three distinct categories:\\
\emph{Genuine shift:} Words that have undergone a clear and significant change in meaning over time.
An example provided in the paper is the word `headed,' which shifted from primarily referring to the `top of a body/entity' to referring to `a direction of travel.'\\
\emph{Borderline:} These are cases where the shift is not as clear-cut, often due to global genre or discourse shifts.
These shifts might be more subtle and may not represent a complete change in meaning but rather a shift in usage within specific contexts.\\
\emph{Corpus artifacts:} These are shifts that arise not from a true change in meaning but rather from peculiarities or biases in the corpus itself.
For example, words like `special,' `cover,' and `romance' might appear to shift in meaning due to their frequent use in certain contexts like book covers or advertisements, which do not necessarily indicate a genuine semantic change.

@@

\begin{table}[tbh]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}}                                          & \multicolumn{3}{c}{\cellcolor[HTML]{FFFFFF}\textbf{Examples}}                                                                                                                                                                                                                               \\ \cmidrule(l){2-4}
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\multirow{-2}{*}{\cellcolor[HTML]{FFFFFF}\textbf{Semantic Shift}}} & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{PPMI}}                                                    & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SVD}}                  & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SGNS}}                                        \\ \midrule
\rowcolor[HTML]{EFEFEF}
\textbf{Genuine Shift}                                                                & started                                                                                                      & \begin{tabular}[c]{@{}l@{}}headed, calls,\\ gay, actually\end{tabular}    & \begin{tabular}[c]{@{}l@{}}wanting, gay, check, starting,\\ major, actually, headed\end{tabular} \\
\cellcolor[HTML]{FFFFFF}\textbf{Borderline}                                           & \begin{tabular}[c]{@{}l@{}}know, got, would, \\ decided, think, stop, \\ remember, must, wanted\end{tabular} & male, naturally                                                           & touching                                                                                         \\
\rowcolor[HTML]{EFEFEF}
\textbf{Corpus Artifact}                                                              & -                                                                                                             & \begin{tabular}[c]{@{}l@{}}harry, wherever,\\ special, cover\end{tabular} & harry, romance                                                                                   \\ \bottomrule
\end{tabular}
\caption{The top 10 words identified by each embedding method were examined, and after reviewing the literature and analyzing their nearest neighbors,
taken from \cite{hamilton-etal-2016-diachronic}.}
\label{tab:discovering-shifts-table}
\end{table}

### 2024-08-20 src/sections/paper_kulkarni.tex
changed introduction and minor changes
introduction changes

\begin{figure}[htb]
    \centering
    \vspace{-1em}
    \includegraphics[scale=0.17]{figures/changepoint}
    \vspace*{-0.5cm}
    \caption{Change Point Algorithm (taken from \cite{kulkarni2014statisticallysignificantdetectionlinguistic}).}
    \label{fig:kulkarni-changepoint}
\end{figure}

\para{Change point algorithm.}
After constructing time series data for each word in the corpus—whether using frequency-based, syntactic, or distributional methods (\Cref{subsec:constructing-word-embeddings}), the first step is to normalize the time series data for the word $w$.
After normalization, the algorithm computes a mean shift series, denoted as $K(Z(w))$.
The mean shift is defined as the difference in the mean of a time series at different time points.
The algorithm calculates this shift by segmenting the time series around a pivot point, denoted as time point $j$, and comparing the means of the segments before and after this point.
To assess the significance of the computed mean shifts, the algorithm employs a bootstrapping method.
This involves generating multiple bootstrap samples, denoted as $BS$.
This is done by repeatedly drawing samples from the normalized data $Z(w)$ until the number of samples equals $B$, which is predefined.
For each sample, the algorithm calculates a p-value for each time point $i$.
This p-value indicates how likely it is that the observed change at that time point is significant.
The algorithm identifies potential change points by looking for times $j$ where the Z-Score $Z_j(w)$ is greater than or equal to the threshold $\gamma$.
It then checks the p-values for these change points and selects the one with the smallest p-value, indicating the most significant change.
The algorithm returns the p-value and the estimated change point (ECP).
This information helps understand exactly when the meaning or usage of the word has shifted over time.
\Cref{fig:kulkarni-changepoint} illustrates the important aspects of the algorithm.

>>>>

%\begin{figure}[htb]
%    \centering
%    \vspace{-1em}
%    \includegraphics[scale=0.17]{figures/changepoint}
%    \vspace*{-0.5cm}
%    \caption{Change Point Algorithm (taken from \cite{kulkarni2014statisticallysignificantdetectionlinguistic}).}
%    \label{fig:kulkarni-changepoint}
%\end{figure}

\para{Change point algorithm.}
After constructing time series data for each word in the corpus—whether using frequency-based, syntactic, or distributional methods (\Cref{subsec:constructing-word-embeddings}), the first step is to normalize the time series data for the word $w$.
After normalization, the algorithm computes a mean shift series, denoted as $K(Z(w))$.
The \emph{mean shift} is defined as the difference in the mean of a time series at different time points.
The algorithm calculates this shift by segmenting the time series around a pivot point, denoted as time point $j$, and comparing the means of the segments before and after this point.
To assess the significance of the computed mean shifts, the algorithm employs a \emph{bootstrapping} method.
The algorithm uses bootstrap sampling to ensure that the detected change point is not just due to random noise.
It generates multiple synthetic time series by randomly resampling the original time series, denoted as $BS$.
These resampled sets serve as a reference to evaluate whether the detected shifts in the real data are significant or not.
This is done by repeatedly drawing samples from the normalized data $Z(w)$ until the number of samples equals $B$, which is predefined.
For each sample, the algorithm calculates a p-value for each time point $i$.
This p-value indicates how likely it is that the observed change at that time point is significant.
The algorithm identifies potential change points by looking for times $j$ where the Z-Score $Z_j(w)$ is greater than or equal to the threshold $\gamma$.
It then checks the p-values for these change points and selects the one with the smallest p-value, indicating the most significant change.
The algorithm returns the p-value and the estimated change point (ECP).
This information helps understand exactly when the meaning or usage of the word has shifted over time.
%\Cref{fig:kulkarni-changepoint} illustrates the important aspects of the algorithm.

### 2024-08-20 src/sections/relatedwork.tex
introduction changes

% ------------------------------------------------------------------ THREE PAPERS ------------------------------------------------------------------
\subsection{Selected Papers}\label{subsec:selected-papers}
Several studies from our literature review provide valuable insights for addressing data voids in our analysis.
The first being the work by Kulkarni et al.\ which introduces us to time series construction.
Their distributional methods focuses on finding subtle semantic shits to determine the context where a word in used.
This concept aligns perfectly with our goal of uncovering shifts in word usage when encountering data voids.
Second paper by Hamilton et al.\ introduces us to word emebeddings alignment.
Since we want to compare word vectors from different time periods, vectors should be aligned in same coordinate axes.
After aligning the embeddings for individual time periods, we can use the aligned word vectors to compute the semantic displacement that a word has undergone during a certain time-period.
Next two sections \Cref{sec:paper_kulkarni}, \Cref{sec:paper_hamilton} focusses on the selected papers in detail.

>>>>

% ------------------------------------------------------------------ THREE PAPERS ------------------------------------------------------------------
\subsection{Selected Papers}\label{subsec:selected-papers}
Several studies from our literature review provided valuable insights for addressing data voids in our analysis.
The first was the work by Kulkarni et al.\ which introduced us to time series construction.
Their distributional methods focused on finding subtle semantic shits to determine the context in which a word in used.
This concept aligned perfectly with our goal of uncovering shifts in word usage when encountering data voids.
The second paper by Hamilton et al.\ introduces us to word emebeddings alignment.
Since we wanted to compare word vectors from different time periods, the vectors had to be aligned on the same coordinate axes.
After aligning the embeddings for individual time periods,
we could use the aligned word vectors to compute the semantic displacement that a word has underwent during a certain time-period.
The next two sections \Cref{sec:paper_kulkarni} and \Cref{sec:paper_hamilton} focused on the selected papers in detail.
